---
title: "Data QC, Density Plots for beta and M values, and PCA"
author: "Beryl"
date: "2015-04-01"
output: 
  html_document:
    keep_md: yes
    toc: yes
---


```{r packages}
suppressPackageStartupMessages(library(ggplot2)) # for plotting
suppressPackageStartupMessages(library(pheatmap)) # for heatmap
suppressPackageStartupMessages(library(RColorBrewer)) # for heatmap palette
suppressPackageStartupMessages(library(plyr))  # data aggregation
suppressPackageStartupMessages(library(dplyr)) #data aggregation
suppressPackageStartupMessages(library(reshape))   #data aggregation
suppressPackageStartupMessages(library(knitr)) # present better tables

```

```{r load_data}

load("../../data/GSE48684_raw_filtered.Rdata")
load("../../data/GSE48684_raw_filtered.beta.norm.Rdata")
load("../../data/GSE48684_raw_filtered.beta.norm.cgi.Rdata")
load("../../data/GSE48684_raw_filtered.m.norm.Rdata")
load("../../data/GSE48684_raw_filtered.m.norm.cgi.Rdata")
load("../../data/metadata.Rdata")

```

####Data QC
To inspect the datasets that we have so far, check the number of NAs. 

```{r}
checkNA <- function(df){
  length(which(is.na(df) == T))
}
order_by_group <- metadata$geo_accession[order(metadata$group, 
																							 metadata$colon_region)]


# raw data, probes filtered
#head(raw_data_filter)
str(raw_data_filter)
checkNA(raw_data_filter)
raw_data_filter <- raw_data_filter[, order_by_group]

# normalized data
#head(beta.norm)
str(beta.norm)
checkNA(beta.norm)
beta.norm <- beta.norm[, order_by_group]

#head(beta.norm.CGI)
str(beta.norm.CGI)
checkNA(beta.norm.CGI)
beta.norm.CGI <- beta.norm.CGI[, order_by_group]

#M value transformed
#head(M.norm)
str(M.norm)
checkNA(M.norm)
#head(M.norm.CGI)
str(M.norm.CGI)
checkNA(M.norm.CGI)
M.norm.CGI <- M.norm.CGI[, order_by_group]

# metadata
load("../../data/metadata.Rdata")
head(metadata)
```

#### Density plots
Plot the density of average beta values of the filtered raw data before and after normalization.

```{r}
library(ggplot2)

aggregateAvgValue <- function(df, cat, group){
	# df is the matrix
	# cat is the beta/M value category (character)
	# group is the sample group (character)
	group_list <- as.character(metadata$geo_accession[which(metadata$group == group)])
	x <- df[, which(colnames(df) %in% group_list)]
	return(data.frame(avg_value = rowMeans(x, na.rm = T),
	                        category = cat,
													 group = group))
}

beta_means <-rbind(
			aggregateAvgValue(raw_data_filter, "raw_beta_value", "normal-H"),
			aggregateAvgValue(raw_data_filter, "raw_beta_value", "normal-C"),
			aggregateAvgValue(raw_data_filter, "raw_beta_value", "cancer"),
			aggregateAvgValue(raw_data_filter, "raw_beta_value", "adenoma"),
			aggregateAvgValue(beta.norm, "normalized_beta_value", "normal-H"),
			aggregateAvgValue(beta.norm, "normalized_beta_value", "normal-C"),
			aggregateAvgValue(beta.norm, "normalized_beta_value", "cancer"),
			aggregateAvgValue(beta.norm, "normalized_beta_value", "adenoma"))

head(beta_means)
str(beta_means)

```

```{r beta_value_density}
ggplot(data = beta_means, aes(x = avg_value, col = category)) +
   geom_density() + 
	xlab("average beta value") +
   ggtitle("Average Beta value density before and after normalization") + 
   theme_bw()


```

#### put this on the poster :)

```{r beta_value_density_facet_by_category}
ggplot(data = beta_means, aes(x = avg_value, col = group)) +
   geom_density() + 
		xlab("average beta value") +
   ggtitle("Average Beta value density distribution") + 
   theme_bw() +
	facet_grid(category ~ .)

# save a large figure for poster
p <- ggplot(data = beta_means, aes(x = avg_value, col = group)) +
   geom_density() + 
		xlab("average beta value") +
   ggtitle("Average Beta value density distribution") + 
   theme_bw() +
	facet_grid(category ~ .) +
	theme(text = element_text(size=28))
ggsave(plot = p, filename = "../../figures/dataQC_beta_density.png", width = 16, height = 10.67, units = "in")
```

Aggregate for M values and plot the density for the 4 groups:

```{r M_value_density}
m_means <-rbind(
			aggregateAvgValue(M.norm, "M_value", "normal-H"),
			aggregateAvgValue(M.norm, "M_value", "normal-C"),
			aggregateAvgValue(M.norm, "M_value", "cancer"),
			aggregateAvgValue(M.norm, "M_value", "adenoma"))

ggplot(data = m_means, aes(x = avg_value)) +
   geom_density() + 
		xlab("average M value") +
   ggtitle("Average M value density \nafter normalization") + 
   theme_bw()
```


```{r M_value_density_facet}

ggplot(data = m_means, aes(x = avg_value, col = group)) +
   geom_density() + 
		xlab("average M value") +
   ggtitle("Average M value density \nafter normalization") + 
   theme_bw()
```


The before and after normalization beta values have similar distribution, and consistent among groups.

The M value distributions have three peaks. The distributions are consistent among groups.

#### Heatmaps

```{r heatmap}
rownames(metadata) <- metadata$geo_accession

# make a function to produce heatmap
plotHeatmap <- function(x, title = "", legend = "group", size = 2, names = F){
  ## x is an ordered matrix
  ## title is the plot title
  ## size is the font size for the rows
  #get color palette
  # pallette code is modified from seminar03, courtesy of Dean Attali
  colour_scheme <- "BuPu"
  palette <- colorRampPalette(rev(brewer.pal(n = 9, colour_scheme)))
  paletteSize <- 256
  cols <- palette(paletteSize)
  #heatmap
  annotation <- metadata[legend] #get the legend
  pheatmap(x, color = cols,
           cluster_rows = FALSE, cluster_cols = FALSE, # turn off dendrogram
           annotation = annotation,
           fontsize_row = size,
           fontsize_col = size,
  				 show_rownames = names,
  				 show_colnames = names,
  				 main = title)
}

#reorder columns by group
order_by_group<-metadata$geo_accession[order(metadata$group, metadata$colon_region)]
beta_cor <- cor(na.omit(raw_data_filter[, order_by_group]))
```

```{r beta_cor_heatmap}
plotHeatmap(beta_cor, "heatmap for the sample correlation\n raw beta value",
						legend = c("group", "colon_region"))
```

```{r beta_norm_beta_cor_heatmap}
norm_beta_cor <- cor(na.omit(beta.norm[, order_by_group]))
plotHeatmap(norm_beta_cor, "heatmap for the sample correlation\n normalized beta value", legend = c("group", "colon_region"))

```

```{r M_cor_heatmap}
M_cor <- cor(na.omit(M.norm[, order_by_group]))
plotHeatmap(M_cor, "heatmap for the sample correlation\n M value", legend = c("group", "colon_region"))

```


#### PCA analysis

PCA for the normalized beta value

```{r pca_normalized_beta_1}

df <- na.omit(beta.norm)
df_pca <- prcomp(df, center = F, scale = F)
plot(df_pca, main = "PCA of normalized beta value")

save(df_pca, file="../../data/pca/beta_norm_pca.Rdata")
df_prin_comp <- cbind(metadata, df_pca$rotation)
save(df_prin_comp, file="../../data/pca/beta_norm_pca_comp.Rdata")
```

```{r pca_normalized_beta}
load("../../data/pca/beta_norm_pca_comp.Rdata")
# scatter plot
ggplot(df_prin_comp, aes(PC1, PC2, label = geo_accession, color = group)) +
  geom_point() +
  ggtitle("Scatterplot of the first two principal components\nnormalized beta value")

# save high res for figures
p <- ggplot(df_prin_comp, aes(PC1, PC2, label = geo_accession, color = group)) +
  geom_point(size = 6) +
	theme(text = element_text(size=28)) +
  ggtitle("Scatterplot of the first two principal components\nnormalized beta value")

ggsave(plot = p, filename = "../../figures/dataQC_norm_beta_PCA.png", width = 16, height = 10.67, units = "in")

```


For M value

```{r pca_M_value_1}

df <- na.omit(M.norm)
df_pca <- prcomp(df, center = F, scale = F)

plot(df_pca, main = "PCA of normalized M value")
df_prin_comp <- cbind(metadata, df_pca$rotation)
save(df_prin_comp, file="../../data/pca/M_norm_pca_comp.Rdata")

```

```{r pca_M_value}
load("../../data/pca/M_norm_pca_comp.Rdata")
# scatter plot
ggplot(df_prin_comp, aes(PC1, PC2, label = geo_accession, color = group)) +
  geom_point() +
  ggtitle("Scatterplot of the first two principal components\nM value")

```



```{r pca_beta_CGI_value_1}

df <- na.omit(beta.norm.CGI)
df_pca <- prcomp(df, center = F, scale = F)

plot(df_pca, main = "PCA of normalized beta CGI value")
df_prin_comp <- cbind(metadata, df_pca$rotation)
save(df_prin_comp, file="../../data/pca/beta_norm_cgi_pca_comp.Rdata")

```

```{r pca_beta_cgi_value}
load("../../data/pca/beta_norm_cgi_pca_comp.Rdata")
# scatter plot
ggplot(df_prin_comp, aes(PC1, PC2, label = geo_accession, color = group)) +
  geom_point() +
  ggtitle("Scatterplot of the first two principal components\nbeta value CGI")

```


```{r pca_M_CGI_value_1}

df <- na.omit(M.norm.CGI)
df_pca <- prcomp(df, center = F, scale = F)
plot(df_pca, main = "PCA of M CGI value")
df_prin_comp <- cbind(metadata, df_pca$rotation)
save(df_prin_comp, file="../../data/pca/M_norm_cgi_pca_comp.Rdata")

```

```{r pca_M_cgi_value}
load("../../data/pca/M_norm_cgi_pca_comp.Rdata")
# scatter plot
ggplot(df_prin_comp, aes(PC1, PC2, label = geo_accession, color = group)) +
  geom_point() +
  ggtitle("Scatterplot of the first two principal components\nM value CGI")

```

