---
title: "Identifying DMR using limma - Part 2"
author: "Eva Y"
date: "April 5, 2015"
output:
  html_document:
    keep_md: yes
---
### Analysis goal: 
1. Get DMR between normal samples (including normal healthy and normal cancer) and adenoma and colorectal cancer samples. 

2. Get DMR between adenoma and colorectal cancer samples. 

### Step 1: Attach packages for analysis
```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(gplots)
library(limma)
library(RColorBrewer)
library(reshape2)
library(VennDiagram)
```


### Step 2: Useful functions
```{r}
# 01 rename sample labels
rename_samples <- function(data, metadata){
  paste(metadata$group, gsub("GSM", "", colnames(data)), sep="_")
}

# 02 make methylation data with only normal-H and normal-C/adenoma/cancer samples
prep_data <- function(data, metadata, sample1="", sample2="", 
                      sample3="", three.samples=FALSE){
  if (three.samples){
  d <- t(data)
  d <- data.frame(sample.id=rep(row.names(d)), 
                                metadata$group, d)
  d <- subset(d, 
              d$metadata.group %in% 
                c(sample1, sample2, sample3))
  
  d <- d[ ,-(2), drop=FALSE]
  row.names(d) <- d$sample.id
  d <- d[ ,-(1), drop=FALSE]
  d <- t(d)
  d  
  } else {
  d <- t(data)
  d <- data.frame(sample.id=rep(row.names(d)), 
                                metadata$group, d)
  d <- subset(d, 
              d$metadata.group %in% 
                c(sample1, sample2))
  
  d <- d[ ,-(2), drop=FALSE]
  row.names(d) <- d$sample.id
  d <- d[ ,-(1), drop=FALSE]
  d <- t(d)
  d
  }
}

# 03 make metadata
prep_metadata <- function(metadata, sample1="", sample2="", sample3="", 
                          level1="", level2="", level3="", regexp="", 
                          three.samples=FALSE, three.levels=FALSE){
  
  if (three.samples){
    metadata_subset <- subset(metadata, 
                               group %in% 
                                 c(sample1, sample2, sample3))
    metadata_subset$group <- gsub(regexp, "", metadata_subset$group)
    } else {
      metadata_subset <- subset(metadata, 
                               group %in% 
                                 c(sample1, sample2))
      metadata_subset$group <- gsub(regexp, "", metadata_subset$group)
    }
  
  if (three.levels){
    metadata_subset$group <- factor(metadata_subset$group, 
                                      levels=c(level1, level2, level3))
    } else {
      metadata_subset$group <- factor(metadata_subset$group, 
                                      levels=c(level1, level2))
    }
  metadata_subset
  }

# 04 perform limma DMA
do.limma <- function(data, design){
  # fit linear model
  fit <- lmFit(data, design)
  
  # apply eBayes() to moderate the estimated error variances
  ebfit <- eBayes(fit)
  
  # get output of the linear model
  topTable(ebfit, number=Inf)
}

# 05 prepare design matrix for different sample pairs
prep_design <- function(metadata="", sample1="", sample2=""){
  metadata$group <- factor(metadata$group, levels=c(sample1, sample2))
  model.matrix(~group, metadata)
}

# 06 plot heatmap for DMR
plot_heatmap <- function(data, palette, labels, metadata, x=""){
  heatmap.2(as.matrix(data), col=palette, 
          trace="none", labRow=NA, labCol=NA,
          ColSideColors=labels[unclass(metadata$group)],
          main=x,
          margin=c(4,10))
  
  # add legend
  legend("topright", legend = levels(metadata$group),
       col=labels, ncol=1, lty=1, lwd=5, cex=0.5)
}

# 07 when given a list of cgi, returns a tall&skinny data frame
get_cgi <- function(data, metadata, int.list, bor.list){
  # get interesting hits
  int <- subset(data, row.names(data) %in% int.list)
  
  # get boring hits
  bor <- subset(data, row.names(data) %in% bor.list)
  
  # make tall and skinny data frame
  df <- rbind(int, bor)
  
  colnames(df) <- rename_samples(df, metadata) # rename colnames
  
  df <- data.frame(cgi=rep(row.names(df)), 
              status=rep(c("interesting", "boring"), each=3), df)
  df <- melt(df, id.vars=c("cgi", "status"), 
             variable.name="sample", value.name="MValue")
  df <- data.frame(df, group=df$sample)
  
  # get rid of "_\\d"
  df$group <- sub("[_]\\d+", "", df$group)
  df
}
```


### Step 3: Load and explore data
```{r}
# load m values and remove NAs
load("../../data/GSE48684_raw_filtered.m.norm.cgi.Rdata")
str(M.norm.CGI, max.level=0)
M.norm.CGI.rmna <- M.norm.CGI[complete.cases(M.norm.CGI), ]
str(M.norm.CGI.rmna, max.level=0)

# save this
save(M.norm.CGI.rmna, file="M.norm.CGI.rmna.Rdata")

# load beta values
load("../03kmeans_cgi/beta.norm.cgi.rmna.Rdata")
str(norm_beta_filter_cgi, max.level=0)

# load metadata
load("../../data/metadata.Rdata")
str(metadata, max.level=0)
```


### Step 4: Data preparation
We want to get DMR between normal samples and adenoma/colorectal cancer samples. Let's prepare the datasets for that.
```{r}
## normal vs. cancer
# make methylation data with only normal and cancer samples
norm_cancer <- prep_data(M.norm.CGI.rmna, metadata, sample1="normal-H", 
               sample2="normal-C", sample3="cancer", three.samples=TRUE)

# make metadata with only normal and cancer samples
metadata_norm_cancer <- prep_metadata(metadata, sample1="normal-H", 
                                      sample2="normal-C", sample3="cancer", 
                                      level1="normal", level2="cancer",
                                      regexp="[-].", 
                                      three.samples=TRUE, 
                                      three.levels=FALSE)

## normal vs. adenoma
# make methylation data with only normal and adenoma samples
norm_adenoma <- prep_data(M.norm.CGI.rmna, metadata, sample1="normal-H", 
               sample2="normal-C", sample3="adenoma", three.samples=TRUE)

# make metadata with only normal and adenoma samples
metadata_norm_adenoma <- prep_metadata(metadata, sample1="normal-H", 
                                      sample2="normal-C", sample3="adenoma", 
                                      level1="normal", level2="adenoma",
                                      regexp="[-].", 
                                      three.samples=TRUE,
                                      three.levels=FALSE)

## adenoma vs. cancer
# make methylation data with only adenoma and cancer samples
adenoma_cancer <- prep_data(M.norm.CGI.rmna, metadata, sample1="adenoma", 
                            sample2="cancer", three.samples=FALSE)

# make metadata with only adenoma and cancer samples
metadata_adenoma_cancer <- prep_metadata(metadata, sample1="adenoma", 
                                         sample2="cancer",
                                         level1="adenoma", 
                                         level2="cancer",
                                         regexp="", 
                                         three.samples=FALSE,
                                         three.levels=FALSE)
```


### Step 5: Perform DMA using `limma`

Make design matrix for different sample pairs.
```{r}
# normal vs. cancer
design_norm_cancer <- prep_design(metadata=metadata_norm_cancer, 
                                   sample1="normal", sample2="cancer")

# normal vs. adenoma
design_norm_adenoma <- prep_design(metadata=metadata_norm_adenoma, 
                                   sample1="normal", sample2="adenoma")

# adenoma vs. cancer
design_adenoma_cancer <- prep_design(metadata=metadata_adenoma_cancer, 
                                   sample1="adenoma", sample2="cancer")
```

Perform DMA using `limma`.
```{r, message=FALSE}
# normal vs. cancer
norm_cancer_dma <- do.limma(norm_cancer, design_norm_cancer)

# normal vs. adenoma
norm_adenoma_dma <- do.limma(norm_adenoma, design_norm_adenoma)

# adenoma vs. cancer
adenoma_cancer_dma <- do.limma(adenoma_cancer, design_adenoma_cancer)

# save DMA results
save(norm_cancer_dma, file="norm_cancer_dma.Rdata")
save(norm_adenoma_dma, file="norm_adenoma_dma.Rdata")
save(adenoma_cancer_dma, file="adenoma_cancer_dma.Rdata")
```


### Step 6: Pick differentially methylated regions at FDR < 1e-05
```{r}
# how many DMR are there at FDR < 1e-05?
norm_cancer_dmr <- subset(norm_cancer_dma, adj.P.Val < 1e-05)
nrow(norm_cancer_dmr)

norm_adenoma_dmr <- subset(norm_adenoma_dma, adj.P.Val < 1e-05)
nrow(norm_adenoma_dmr)

adenoma_cancer_dmr <- subset(adenoma_cancer_dma, adj.P.Val < 1e-05)
nrow(adenoma_cancer_dmr)
```


### Step 7: Explore DMA and visualize DMR

What are the shared DMR between these different sample pairs analyses?
```{r}
# assign names to the lists for plotting
dmr <- list(Normal.Adenoma=row.names(norm_adenoma_dmr), 
            Normal.Cancer=row.names(norm_cancer_dmr), 
            Adenoma.Cancer=row.names(adenoma_cancer_dmr))

# start a new plot
plot.new()

# draw the Venn diagram
venn.plot <- venn.diagram(dmr, filename = NULL, 
                          fill = c("red", "blue", "green"))

# draw the plot on the screen
grid.draw(venn.plot)
```

Plot heatmap for each DMR list.
```{r}
# color palette 
jRdBu <- colorRampPalette(brewer.pal(n=9, "RdBu"))
palette.size <- 256
jRdBu.palette <- jRdBu(palette.size)

# specify color labels for different samples
sample.cancer <- brewer.pal(11, "Spectral")[c(3,1)]
sample.cancer.3 <- brewer.pal(11, "Spectral")[c(3,9,1)]

sample.adenoma <- brewer.pal(11, "Spectral")[c(3,5)]
sample.adenoma.3 <- brewer.pal(11, "Spectral")[c(3,9,5)]

sample.cancer.adenoma <- brewer.pal(11, "Spectral")[c(5,1)]

# create data frame for heatmap.2
# norm vs. cancer
hmap_normcancer <- subset(norm_cancer, 
                           row.names(norm_cancer) %in% 
                             row.names(norm_cancer_dmr))
hmap_normcancer <- hmap_normcancer[1:450, ]

# norm vs. adenoma
hmap_normadenoma <- subset(norm_adenoma, 
                           row.names(norm_adenoma) %in% 
                             row.names(norm_adenoma_dmr))
hmap_normadenoma <- hmap_normadenoma[1:450, ]

# adenoma vs. cancer
hmap_adenomacancer <- subset(adenoma_cancer, 
                           row.names(adenoma_cancer) %in% 
                             row.names(adenoma_cancer_dmr))
hmap_adenomacancer <- hmap_adenomacancer[1:450, ]

# make new metadata that discrimates between normal-H and normal-C
m.norm.cancer.3 <- prep_metadata(metadata, sample1="normal-H", 
                   sample2="normal-C", sample3="cancer",
                   level1="normal-H", level2="normal-C",
                   level3="cancer",
                   regexp="", three.samples=TRUE, 
                   three.levels=TRUE)

m.norm.adenoma.3 <- prep_metadata(metadata, sample1="normal-H", 
                   sample2="normal-C", sample3="adenoma",
                   level1="normal-H", level2="normal-C",
                   level3="adenoma",
                   regexp="", three.samples=TRUE, 
                   three.levels=TRUE)

# plot heatmaps
# normal vs. cancer

png("hmap_norm_cancer.png", res = 300, 
    width = 5*300, height = 5*300,
    pointsize = 8)

plot_heatmap(hmap_normcancer, jRdBu.palette, sample.cancer, 
             metadata_norm_cancer, x="Normal vs. Cancer")

dev.off()

png("hmap_norm_cancer3.png", res = 300, 
    width = 5*300, height = 5*300,
    pointsize = 8)

plot_heatmap(hmap_normcancer, jRdBu.palette, sample.cancer.3, 
             m.norm.cancer.3, x="Normal vs. Cancer")

dev.off()

# normal vs. adenoma

png("hmap_norm_adenoma.png", res = 300, 
    width = 5*300, height = 5*300,
    pointsize = 8)

plot_heatmap(hmap_normadenoma, jRdBu.palette, sample.adenoma, 
             metadata_norm_adenoma, x="Normal vs. Adenoma")

dev.off()

png("hmap_norm_adenoma3.png", res = 300, 
    width = 5*300, height = 5*300,
    pointsize = 8)

plot_heatmap(hmap_normadenoma, jRdBu.palette, sample.adenoma.3, 
             m.norm.adenoma.3, x="Normal vs. Adenoma")

dev.off()

# adenoma vs. cancer

png("hmap_adenoma_cancer.png", res = 300, 
    width = 5*300, height = 5*300,
    pointsize = 8)

plot_heatmap(hmap_adenomacancer, jRdBu.palette, sample.cancer.adenoma, 
             metadata_adenoma_cancer, x="Adenoma vs. Cancer")

dev.off()
```


Plot some interesting and boring hits for each DMA. 
```{r}
# normal vs. cancer
int.list <- head(row.names(norm_cancer_dma), 3)
bor.list <- tail(row.names(norm_cancer_dma), 3)
norm_cancer_cgi <- get_cgi(norm_cancer, metadata_norm_cancer, int.list, bor.list)

# plot
ggplot(norm_cancer_cgi, aes(x=group, y=MValue, color=status)) +
  geom_point() + 
  facet_wrap(~cgi) +
  theme_bw() +
  stat_summary(aes(group=1), fun.y=mean, geom="line") +
  ggtitle("Normal vs. Cancer")

ggsave(file="norm_cancer_cgi.png", dpi=300)

# normal vs. adenoma
int.list <- head(row.names(norm_adenoma_dma), 3)
bor.list <- tail(row.names(norm_adenoma_dma), 3)
norm_adenoma_cgi <- get_cgi(norm_adenoma, metadata_norm_adenoma, int.list, bor.list)

# plot
ggplot(norm_adenoma_cgi, aes(x=group, y=MValue, color=status)) +
  geom_point() + 
  facet_wrap(~cgi) +
  theme_bw() +
  stat_summary(aes(group=1), fun.y=mean, geom="line") +
  ggtitle("Normal vs. Adenoma")

ggsave(file="norm_adenoma_cgi.png", dpi=300)

# adenoma vs. cancer
int.list <- head(row.names(adenoma_cancer_dma), 3)
bor.list <- tail(row.names(adenoma_cancer_dma), 3)
adenoma_cancer_cgi <- get_cgi(adenoma_cancer, metadata_adenoma_cancer, int.list, bor.list)

# plot
ggplot(adenoma_cancer_cgi, aes(x=group, y=MValue, color=status)) +
  geom_point() + 
  facet_wrap(~cgi) +
  theme_bw() +
  stat_summary(aes(group=1), fun.y=mean, geom="line") +
  ggtitle("Adenoma vs. Cancer")

ggsave(file="adenoma_cancer_cgi.png", dpi=300)
```

## Summary
Guys, this looks good!
