---
title: 'Two Group: Normal and Cancer'
author: "Santina"
date: "Sunday, April 05, 2015"
output: 
  html_document:
    keep_md: yes
    toc: yes
---

# Goal 
Beryl and Eva have been trying to do limma on normal samples (normal-H + normal-C) and cancer samples. They are getting different results so I'll try to see if I can reproduce one of their results. 



# Subset Data 

I'll use normalized cgi M value. 

Load data: 
```{r}
load("../../data/GSE48684_raw_filtered.m.norm.cgi.Rdata") # M.norm.CGI
load('../../data/metadata.Rdata') # metadata
````

Inspect 
```{r}
head(M.norm.CGI[, 1:5])
head(metadata)

#review what we have in our samples 
table(metadata$group)
```


Subset 

We'll use metadata to know which columns in the `M.norm.CGI` to keep.
```{r}
# sample_names <- rownames(metadata[metadata$group %in% c("normal-H", "normal-C", "cancer"), ]) 
# data_sub <- M.norm.CGI[, sample_names]
# I won't do this so I can keep things in order 

# get column names for "normal"
normal_samples <- rownames(metadata[metadata$group %in% c("normal-H", "normal-C"), ])

# get column names for "cancer"
cancer_samples <- rownames(metadata[metadata$group == "cancer", ])

normal_data <- M.norm.CGI[, normal_samples]
ncol(normal_data)  # 24 + 17 = 41

cancer_data <- M.norm.CGI[, cancer_samples]
ncol(cancer_data)  

data_sub <- cbind(normal_data, cancer_data)
ncol(data_sub) # instead of 147
````

Data cleaning: removing rows with NA values 
```{r}
(a <- nrow(data_sub))
data_sub <- data_sub[complete.cases(data_sub), ]
(a1 <- nrow(data_sub))
```

So we have removed `r a-a1` rows.

# Limma 

Time to do the real stuff ! 

Design matrix: 
```{r}
des <- data.frame(replicate = colnames(data_sub), 
                                    condition = factor(c(rep("normal", 41), rep("cancer", 64)))) # we can do this because we know the colnames are in order 


des <- model.matrix(~condition, des)
head(des)
```


```{r}
library(limma)

limmaFit <- lmFit(data_sub, des)
EbFit <- eBayes(limmaFit)
EbFit_table <- topTable(EbFit, number = nrow(EbFit))
head(EbFit_table)
```


Just arrange the table a bit
```{r}
# keep  needed columns 
EbFit_table <- EbFit_table[, c("logFC", "t", "P.Value", "adj.P.Val")]
# rename column names 
colnames(EbFit_table)  <- c("log.fc", "test.stat", "p.value", "q.value")
head(EbFit_table)
```

Looks good, now I'll save it : 

```{r}
write.table(EbFit_table, file="normal_vs_cancer_santina.tsv", sep = "\t")
```



